Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2024-01-05 E-Hypergraphs-LICS24.tex
Sections 3.2 and 5

\end{definition}
When $\mathbb{C} = \catname{Hyp_{\Sigma}}$, the rewrite rules become cospans of hypergraphs with matching interfaces: $i \xrightarrow{} L \xleftarrow{} j$, $i \xrightarrow{} R \xleftarrow{} j$.
By noting that a cospan $0 \xrightarrow{} L \xleftarrow{[f_{i},f_{j}]} i + j$ encodes the same data as a cospan $i \xrightarrow{f_{i}} L \xleftarrow{f_{j}} j$ (Remark 3.14~\cite{Frobenius}) we can formulate a rewrite rule as a single span $L \xleftarrow{} K \xrightarrow{} R$ where $K \cong i + j$.
Likewise, for the target cospan $n \xrightarrow{} G \xleftarrow{} m$  $J \cong n + m$.
% However, this notion of DPO does not account for hypergraphs with interfaces, i.e. those from $\MdaCospans$ where $l,r$ from $\langle l, r \rangle$ are cospans. To support this we first assume that we only need to maintain the interfaces of $i \xrightarrow{} L \xleftarrow{} j$ and $i \xrightarrow{} R \xleftarrow{} j$ during rewriting, then we can take $K \cong i + j$, i.e. take $K$ to be the coproduct of $i,j$ (remark 3.14~\cite{Frobenius}). What remains is to account for the interfaces in $G$, $C$, and $H$, which brings us to the following modified DPOI square. $J \cong n + m$ where $n$ and $m$ are the input and output interfaces of $G$, $C$, and $H$ respectively. The only difference from DPO square above is that we now explicitly require that the initial object $G$, intermediate object $C$ and final object $H$ all share the same interfaces.

@@

\begin{definition}
Given $G \xleftarrow{} n+m$ and $H \xleftarrow{} n + m$ in $\catname{Hyp_{\Sigma}}$, $G$ rewrites convexly into $H$ with interface $n + m$ --- notation $(G \xleftarrow{} n + m ) \Rrightarrow_{\mathcal{R}}  (H \xleftarrow{} n + m )$ --- if there exist rule $L \xleftarrow{} i + j \xrightarrow{} R$ in $\mathcal{R}$ and object $C$ and cospan arrows $i+j \xrightarrow{} C \xleftarrow{} n+m$ such that the DPOI diagram above commutes and its marked squares are pushouts and the following conditions hold

@@

@@

We have a more general notion of interface and hence we need to adjust DPOI rewriting a little bit.

We have a more general notion of interface and hence we need to adjust DPOI rewriting a little bit.
The first issue is that whole interfaces are not preserved during rewriting, consider an example in Figure~\ref{fig:example_rewrite_ecospans} which corresponds to a rewrite rule of $\langle f, f+g \rangle$.

%\begin{figure}[t]
%    \centering
%    \[
%    \scalebox{0.6}{\tikzfig{combinatorial_semantics/f_to_f_plus_g_example}}
%    \]
%    \caption{Example rewrite rule in $\WellTypedMdaEcospans$}
%    \label{fig:example_rewrite_ecospans}
%\end{figure}

The input interface on the left-hand side of the rule is $\{1, 2\}$, while the input interface on the right-hand side of the rule is $\{1,2,5,6,7,8\}$.
Note that the outermost interfaces are preserved.

Another concern is when we delete the occurrence of the left-hand side of a rewrite rule from an e-hypergraph we also modify this e-hypergraph's interfaces.
Consider a hypothetical DPO square below in Figure~\ref{fig:interface_change_example} where long arrows denote matching between carriers of cospans.

The first row of the diagram shows the span for a rewrite rule of $\langle f + g, f \rangle$ which is like a projection rule.
Note once again the difference between the interfaces of the left-hand and right-hand sides of the rewrite rule.
The entry in the middle of the span depicts the outermost interfaces which are preserved.
The second row depicts the e-hypergraph to be rewritten on the left and the result of the rewrite on the right.
The result is obtained by first deleting the image of the left-hand side of the rewrite rule from the initial e-hypergraph (depicted in the middle) and then gluing the right-hand side of the rewrite rule into it. 
Note that the interfaces of the initial e-hypergraph and the residual e-hypergraph are different. 
Here, the outermost and the whole interfaces for the eventual e-hypergraph coincide, but if the nested-ness of the initial e-hypergraph was more intense the result could have inner interfaces as well as we would remove the interfaces of the inner edges when deleting the image of the left-hand side part of the rule from the initial e-hypergraph.

\update{$i,j$ are replaced}

To account for such changes in the interfaces we put some additional constraints on DPOI rewriting.
Consider a DPOI diagram with extra constraints below which encodes
 the notion that cospan 
 $n_{ext} \xrightarrow{} g_i \xrightarrow{} G \xleftarrow{} g_o \xleftarrow{} m_{ext}$
 rewrites into a cospan 
 $n_{ext} \xrightarrow{} h_i \xrightarrow{} H \xleftarrow{} h_o \xleftarrow{} m_{ext}$ via a rewrite rule $\langle l, r \rangle$,
 where
 $l := x_{ext} \xrightarrow{} l_i \xrightarrow{} L \xleftarrow{} l_o \xleftarrow{} y_{ext}$
 and
 $r := x_{ext} \xrightarrow{} r_i \xrightarrow{} L \xleftarrow{} r_o \xleftarrow{} y_{ext}$.
 Note there are arrows from $x_{ext} + y_{ext} \to l_i + l_o$ (respectively, $r_i + r_o$) from the corresponding cospans which are not shown in the diagram.
 Similarly for the arrows from $n_{ext} + m_{ext}$.

% \[\begin{tikzcd}
% 	&&&& {i + j} \\
% 	{l_i+l_o} && L &&&& R && {r_i+r_o} \\
% 	\\
% 	{g_i+g_o} && {G^{\urcorner}} && C && {^{\ulcorner}H} && {h_i+h_o} \\
% 	\\
% 	&&&& {n+m}
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=1-5, to=2-3]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=1-5, to=2-7]
% 	\arrow["m", color={rgb,255:red,214;green,92;blue,92}, from=2-3, to=4-3]
% 	\arrow["{[i_c,j_c]}", color={rgb,255:red,214;green,92;blue,92}, from=1-5, to=4-5]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=2-7, to=4-7]
% 	\arrow["{[n_c,m_c]}"', color={rgb,255:red,214;green,92;blue,92}, from=6-5, to=4-5]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=4-5, to=4-7]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=4-5, to=4-3]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=6-5, to=4-3]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=6-5, to=4-7]
% 	\arrow[from=2-1, to=2-3]
% 	\arrow["{[f_i,f_o]}", from=2-1, to=4-1]
% 	\arrow[from=4-1, to=4-3]
% 	\arrow[from=2-9, to=2-7]
% 	\arrow[from=4-9, to=4-7]
% 	\arrow["{[h_i,h_o]}"{description}, from=2-9, to=4-9]
% 	\arrow[from=6-5, to=4-9]
% 	\arrow[from=6-5, to=4-1]
% 	\arrow["{[i_l,j_l]}"{description}, from=1-5, to=2-1]
% 	\arrow["{[i_r,j_r]}"{description}, from=1-5, to=2-9]
% \end{tikzcd}\]

@@

The parts of the original DPOI square is shown in blue.
The extra bits simply account for more complicated interfaces.
For example, the left-hand side $L$ of a rewrite rule
 $\langle L, R \rangle$ which originally was some cospan
 $i \xrightarrow{} L \xleftarrow{} j$ in $\MdaCospans$ 
 now is a cospan
 $x_{ext} \xrightarrow{} l_i \to L \xleftarrow{} l_o \xleftarrow{} y_{ext}$
from $\WellTypedMdaEcospans$ where $l_i, l_o$ are whole input and output interfaces respectively.
The top half of this diagram says that the outermost interfaces
 should coincide for $L$ and $R$.
The bottom half puts the same restrictions on $G$ and $H$.
For the pushout to exist we impose our assumptions~\ref{pushout:assumptions} on the correspoding arrow.

We also adjust the definition of a boundary complement.
\begin{definition}[Down-closed subgraph]
We call a subgraph $\mathcal{H}$ of an e-hypergraph $\mathcal{G}$ \emph{down-closed} if for all $e \in E_{\mathcal{H}}$,   every element of $(e]$ is also in $\mathcal{H}$.
\end{definition}

@@

% https://q.uiver.app/#q=WzAsNyxbMiwwLCJMIl0sWzQsMCwieF97ZXh0fSArIHlfe2V4dH0iXSxbMiwyLCJHXntcXHVyY29ybmVyfSJdLFs0LDIsIkMiXSxbNCw0LCJuX3tleHR9K21fe2V4dH0iXSxbMCwwLCJsX2krbF9vIl0sWzAsMiwiZ19pK2dfbyJdLFsxLDAsIiIsMCx7ImNvbG91ciI6WzAsNjAsNjBdfV0sWzAsMiwibSIsMCx7ImNvbG91ciI6WzAsNjAsNjBdfSxbMCw2MCw2MCwxXV0sWzEsMywiW2lfYyxqX2NdIiwwLHsiY29sb3VyIjpbMCw2MCw2MF19LFswLDYwLDYwLDFdXSxbNCwzLCJbbl9jLG1fY10iLDIseyJjb2xvdXIiOlswLDYwLDYwXX0sWzAsNjAsNjAsMV1dLFszLDIsIiIsMix7ImNvbG91ciI6WzAsNjAsNjBdfV0sWzQsMiwiIiwyLHsiY29sb3VyIjpbMCw2MCw2MF19XSxbNSwwXSxbNiwyXV0=
\[
\scalebox{0.75}{
    \tikzfig{combinatorial_semantics/DPOI_pushout_complement}
}
\]

Consider the diagram above where the marked square is a pushout
 and $x_{ext} + y_{ext}$ and $n_{ext} + m_{ext}$ are 
 the \textit{outermost} interfaces for the corresponding 
 cospans.
We say that $x_{ext} + y_{ext} \to C \to G^{\urcorner}$ is 
 a \textit{boundary} complement if $[i_c, i_j]$ is mono 
 and there exist $n_c : n_{ext} \to C$ and $m_c : m_{ext} \to C$ 
 making the above triangle commute and such that
there exists a \textit{not necessarily well-typed} mda-cospan

@@

where $c_i$ and $c_o$ are computed as
 $c_i = g_i \setminus (l_i \setminus x_{ext})$ and
 $c_o = g_o \setminus (l_o \setminus y_{ext})$.
\end{definition}
%
%\begin{figure}
%    \centering
%    \[
%    \scalebox{0.5}{\tikzfig{combinatorial_semantics/interfaces_change_example}}\]
%    \caption{Hypothetical DPO square for $\catname{MACsp_{D}(EHyp_{\Sigma})}$}
%    \label{fig:interface_change_example}
%\end{figure}

@@

\begin{remark}
Note that cospan $n_{ext} \xrightarrow{} c_i + y_{ext} \xrightarrow{} C \xleftarrow{} c_o + x_{ext} \xleftarrow{} m_{ext}$ is not well-typed.
The cospan's input interface contains nodes that previously were a part of the output interface.
\end{remark}

\begin{proposition}[Proposition~\ref{prop:boundary_complement} restatement]
    The boundary complement in~\ref{def:boundary_new} when exists is unique.
\end{proposition}

@@

    \begin{enumerate}
        \item $m$ is monomorphism
        \item $m(L)$ is a convex down-closed subgraph of $G$
        % \item \update{I've removed the requirement that consistency should be reflected because well typedness prohibits otherwise formed matches.}
    \end{enumerate}
\end{definition}
\

\begin{definition}[Rewriting in $\WellTypedMdaEcospans$]
    Given $G \xleftarrow{} g_i + g_o \xleftarrow{} n_{ext} + m_{ext}$ 
    and $H \xleftarrow{} h_i + h_o \xleftarrow{} n_{ext} + m_{ext}$ in $\Ecospans$ with the outermost interfaces $n_{ext} + m_{ext}$, $G$ rewrites convexly into $H$ with the outermost interface $n_{ext} + m_{ext}$ --- notation $( G \xleftarrow{} n_{ext} + m_{ext} ) \Rrightarrow_{\mathcal{R}}^{+} (E \xleftarrow{} n_{ext} + m_{ext} )$ --- if there exists a rule $L \xleftarrow{} l_i + l_o \xleftarrow{} x_{ext} + y_{ext} \xrightarrow{} r_i + r_o \xrightarrow{} R$ with the outermost interfaces $x_{ext} + y_{ext}$ in $\mathcal{R}$ and object $C$ and cospan arrows $x_{ext} + y_{ext} \xrightarrow{} C \xleftarrow{} n_{ext} + m_{ext}$ in $\Ecospans$ such that the diagram below commutes and its marked squares are pushouts

@@

    \begin{enumerate}
        \label{dpoi:assumptions}
        \item $m : L \to G$ is a convex down-closed match;
        \item $[m(v_i)) = [m(v_j))$ for all $v_i,v_j$ that have a pre-image in $x_{ext} + y_{ext}$
        \item $\consistency^{\mu}(m(v_i)) = \consistency^{\mu}(m(v_j))$ for all $v_i,v_j$ that have a pre-image in $x_{ext} + y_{ext}$
        \item $x_{ext} + y_{ext} \to C \to G$ is a boundary complement in the leftmost pushout.
    \end{enumerate}
\end{definition}

\begin{remark}
The conditions 2 and 3 are needed to guarantee the existence of the pushout complement.
\end{remark}

The whole interfaces of the resulting e-hypergraph $H$ from the cospan
 $n_{ext} \xrightarrow{} h_i \xrightarrow{} H \xleftarrow{} h_o \xleftarrow{} m_{ext}$
 are constructed as $h_i = (r_i \setminus x_{ext}) + c_i$ 
 and
 $h_o = (r_o \setminus y_{ext}) + c_o$.
 $h_i \to H$ is defined as a copairing of 
 $f_1 : r_i \setminus x_{ext} \to R \to H$ 
 and 
 $f_2 : c_i \to C \to H$. 
Since both $f_1$ and $f_2$ are mono, their copairing is also a mono. 
Similarly for $h_o$.

@@

if 
\[
    G \xleftarrow{\langle a_2,b_2 \rangle} g_i + g_{o} \xleftarrow{\langle a_1, b_1 \rangle} n_{ext} + m_{ext} \Rrightarrow H \xleftarrow{\langle a_2,b_2 \rangle} h_i + h_{o} \xleftarrow{\langle a_1, b_1 \rangle} n_{ext} + m_{ext}
\]
under a rewrite rule
\[
    \langle l_{i} + l_{o} \xrightarrow{\langle a_2, b_2 \rangle} L \xleftarrow{\langle a_1, b_1 \rangle} x_{ext} + y_{ext},
    r_{i} + r_{o} \xrightarrow{\langle a_2, b_2 \rangle} R \xleftarrow{\langle a_1, b_1 \rangle} x_{ext} + y_{ext} 
    \rangle
\]

@@

\begin{theorem}[Existence of pushouts in $\catname{EHyp_{\Sigma}}$]
\label{th:existence_of_pushouts}
Consider the following span in $\catname{EHyp_{\Sigma}}$
% https://q.uiver.app/#q=WzAsMyxbMCwwLCJaIl0sWzIsMCwiWCJdLFswLDIsIlkiXSxbMCwxLCJmIl0sWzAsMiwiZyIsMl1d
\[\begin{tikzcd}
	Z && X \\
	\\
	Y
	\arrow["f", from=1-1, to=1-3]
	\arrow["g"', from=1-1, to=3-1]
\end{tikzcd}\]
such that
\begin{enumerate}
    \item $Z$ is a \textit{descrete} e-hypergraph.
    \item $f,g$ are monos.
    \item $[f_{V}(v_i)) = [f_{V}(v_j))$ and $[g_{V}(v_i)) = [g_{V}(v_j))$ for all $v_{i},v_{j}$ in $V_{Z}$.
    \item If $[f_{V}(v)) \not = \varnothing$ then $[g_{V}(v)) = \varnothing$ and if $[g_{V}(v)) \not = \varnothing$ then $[f_{V}(v)) = \varnothing$.
    \item $\consistency^{\mu}(f(v_i)) = \consistency^{\mu}(f(v_j))$ and $\consistency^{\mu}(g(v_i)) = \consistency^{\mu}(g(v_j))$ for all $v_i,v_j$ in $V_{Z}$.
\end{enumerate}    
then the pushout $X +_{f,g} Y$ exists.
\end{theorem}
\begin{proof}
    Recall the diagram below.
    \[
        \begin{tikzcd}
        Z \arrow[r, "f"] \arrow[d, "g"']                                   & X \arrow[d, "\sfrac{i_1}{\sim R}"] \arrow[rdd, "j_1", bend left] &   \\
        Y \arrow[r, "\sfrac{i_2}{\sim R}"] \arrow[rrd, "j_2"', bend right] & \sfrac{X+Y}{\sim R} \arrow[rd, "u"]                              &   \\
                                                                        &                                                                  & Q
        \end{tikzcd}
    \]
    Then, the pushout of e-hypergraphs $X$ and $Y$ is computed in two steps.
    First, a coproduct of $X+Y$ is computed which is 
    \[
        X + Y = \{V_{X} + V_{Y}, E_{X} + E_{Y}, s_{X+Y}, t_{X+Y}, \consistency_{X+Y}, <_{c,X+Y} \}
    \]

@@

>>>>

\end{definition}
When $\mathbb{C} = \catname{Hyp_{\Sigma}}$, the rewrite rules become cospans of hypergraphs with matching interfaces: $i \xrightarrow{} L \xleftarrow{} j$, $i \xrightarrow{} R \xleftarrow{} j$.
By noting that a cospan $0 \xrightarrow{} L \xleftarrow{[f_{i},f_{j}]} i + j$ encodes the same data as a cospan $i \xrightarrow{f_{i}} L \xleftarrow{f_{j}} j$ (Remark 3.14~\cite{Frobenius}) we can formulate a rewrite rule as a single span $L \xleftarrow{} K \xrightarrow{} R$ where $K \cong i + j$.
Likewise, for the target cospan $n \xrightarrow{} G \xleftarrow{} m$,  $J \cong n + m$.
% However, this notion of DPO does not account for hypergraphs with interfaces, i.e. those from $\MdaCospans$ where $l,r$ from $\langle l, r \rangle$ are cospans. To support this we first assume that we only need to maintain the interfaces of $i \xrightarrow{} L \xleftarrow{} j$ and $i \xrightarrow{} R \xleftarrow{} j$ during rewriting, then we can take $K \cong i + j$, i.e. take $K$ to be the coproduct of $i,j$ (remark 3.14~\cite{Frobenius}). What remains is to account for the interfaces in $G$, $C$, and $H$, which brings us to the following modified DPOI square. $J \cong n + m$ where $n$ and $m$ are the input and output interfaces of $G$, $C$, and $H$ respectively. The only difference from DPO square above is that we now explicitly require that the initial object $G$, intermediate object $C$ and final object $H$ all share the same interfaces.

@@

\begin{definition}
Given $G \xleftarrow{} n+m$ and $H \xleftarrow{} n + m$ in $\catname{Hyp_{\Sigma}}$, $G$ rewrites convexly into $H$ with interface $n + m$ --- notation $(G \xleftarrow{} n + m ) \Rrightarrow_{\mathcal{R}}  (H \xleftarrow{} n + m )$ --- if there exist rule $L \xleftarrow{} i + j \xrightarrow{} R$ in the set of rewrite rules $\mathcal{R}$ and object $C$ and cospan arrows $i+j \xrightarrow{} C \xleftarrow{} n+m$ such that the DPOI diagram above commutes and its marked squares are pushouts and the following conditions hold

@@

By resorting to convex matching and boundary complement we virtually require that the original cospans above are monogamous, e.g., $n \xrightarrow{} G \xleftarrow{} m$.

@@

As extended cospans have a more general notion of interface we need to make some adjustments to DPOI rewriting.
The main issue is that internal interfaces may not be preserved during rewriting, for example, if the right-hand side of the rewrite rule has more hierarchy than the left-hand side: it affects both the eventual e-hypergraph and the possible pushout complement.

% The input interface on the left-hand side of the rule is $\{1, 2\}$, while the input interface on the right-hand side of the rule is $\{1,2,5,6,7,8\}$.
% Note that the outermost interfaces are preserved.

% Another concern is when we delete the occurrence of the left-hand side of a rewrite rule from an e-hypergraph we also modify this e-hypergraph's interfaces.
% Consider a hypothetical DPO square below in Figure~\ref{fig:interface_change_example} where long arrows denote matching between carriers of cospans.

% The first row of the diagram shows the span for a rewrite rule of $\langle f + g, f \rangle$ which is like a projection rule.
% Note once again the difference between the interfaces of the left-hand and right-hand sides of the rewrite rule.
% The entry in the middle of the span depicts the outermost interfaces which are preserved.
% The second row depicts the e-hypergraph to be rewritten on the left and the result of the rewrite on the right.
% The result is obtained by first deleting the image of the left-hand side of the rewrite rule from the initial e-hypergraph (depicted in the middle) and then gluing the right-hand side of the rewrite rule into it. 
% Note that the interfaces of the initial e-hypergraph and the residual e-hypergraph are different. 
% Here, the outermost and the whole interfaces for the eventual e-hypergraph coincide, but if the nested-ness of the initial e-hypergraph was more intense the result could have inner interfaces as well as we would remove the interfaces of the inner edges when deleting the image of the left-hand side part of the rule from the initial e-hypergraph.

To account for such changes in the interfaces we put some additional constraints on DPOI rewriting for the case of $\catname{EHyp_{\Sigma}}$.
Consider a DPOI diagram with extra constraints below which encodes
the notion that e-hypergraph $G$ rewrites into $H$ with external interfaces $n_{ext}+m_{ext}$.
Note the new interfaces and how we use the same trick to go from cospan $n_{ext} \xrightarrow{} g_i \xrightarrow{} G \xleftarrow{} g_o \xleftarrow{} m_{ext}$ to $G \xleftarrow{} g_{i} + g_{o} \xleftarrow{} n_{ext} + m_{ext}$.
The original DPO square is depicted in blue.
% \[\begin{tikzcd}
% 	&&&& {i + j} \\
% 	{l_i+l_o} && L &&&& R && {r_i+r_o} \\
% 	\\
% 	{g_i+g_o} && {G^{\urcorner}} && C && {^{\ulcorner}H} && {h_i+h_o} \\
% 	\\
% 	&&&& {n+m}
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=1-5, to=2-3]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=1-5, to=2-7]
% 	\arrow["m", color={rgb,255:red,214;green,92;blue,92}, from=2-3, to=4-3]
% 	\arrow["{[i_c,j_c]}", color={rgb,255:red,214;green,92;blue,92}, from=1-5, to=4-5]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=2-7, to=4-7]
% 	\arrow["{[n_c,m_c]}"', color={rgb,255:red,214;green,92;blue,92}, from=6-5, to=4-5]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=4-5, to=4-7]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=4-5, to=4-3]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=6-5, to=4-3]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=6-5, to=4-7]
% 	\arrow[from=2-1, to=2-3]
% 	\arrow["{[f_i,f_o]}", from=2-1, to=4-1]
% 	\arrow[from=4-1, to=4-3]
% 	\arrow[from=2-9, to=2-7]
% 	\arrow[from=4-9, to=4-7]
% 	\arrow["{[h_i,h_o]}"{description}, from=2-9, to=4-9]
% 	\arrow[from=6-5, to=4-9]
% 	\arrow[from=6-5, to=4-1]
% 	\arrow["{[i_l,j_l]}"{description}, from=1-5, to=2-1]
% 	\arrow["{[i_r,j_r]}"{description}, from=1-5, to=2-9]
% \end{tikzcd}\]

@@

% The parts of the original DPOI square is shown in blue.
% For example, the left-hand side $L$ of a rewrite rule
%  $\langle L, R \rangle$ which originally was some cospan
%  $i \xrightarrow{} L \xleftarrow{} j$ in $\MdaCospans$ 
%  now is a cospan
%  $x_{ext} \xrightarrow{} l_i \to L \xleftarrow{} l_o \xleftarrow{} y_{ext}$
% from $\WellTypedMdaEcospans$ where $l_i, l_o$ are whole input and output interfaces respectively.
% The top half of this diagram says that the outermost interfaces
%  should coincide for $L$ and $R$.
% The bottom half puts the same restrictions on $G$ and $H$.
% For the pushout to exist we impose our assumptions~\ref{pushout:assumptions} on the correspoding arrow.

We then also adjust the definition of a boundary complement.

@@

% https://q.uiver.app/#q=WzAsNyxbMiwwLCJMIl0sWzQsMCwieF97ZXh0fSArIHlfe2V4dH0iXSxbMiwyLCJHXntcXHVyY29ybmVyfSJdLFs0LDIsIkMiXSxbNCw0LCJuX3tleHR9K21fe2V4dH0iXSxbMCwwLCJsX2krbF9vIl0sWzAsMiwiZ19pK2dfbyJdLFsxLDAsIiIsMCx7ImNvbG91ciI6WzAsNjAsNjBdfV0sWzAsMiwibSIsMCx7ImNvbG91ciI6WzAsNjAsNjBdfSxbMCw2MCw2MCwxXV0sWzEsMywiW2lfYyxqX2NdIiwwLHsiY29sb3VyIjpbMCw2MCw2MF19LFswLDYwLDYwLDFdXSxbNCwzLCJbbl9jLG1fY10iLDIseyJjb2xvdXIiOlswLDYwLDYwXX0sWzAsNjAsNjAsMV1dLFszLDIsIiIsMix7ImNvbG91ciI6WzAsNjAsNjBdfV0sWzQsMiwiIiwyLHsiY29sb3VyIjpbMCw2MCw2MF19XSxbNSwwXSxbNiwyXV0=

For monogamous extended cospans $x_{ext} \xrightarrow{} l_{i} L \xleftarrow{} l_{o} \xleftarrow{} y_{ext}$ and $n_{ext} \xrightarrow{} g_{i} \xrightarrow{} G \xleftarrow{} g_{o} \xleftarrow{} m_{ext}$ and mono $f : L \to G$, a pushout complement $x_{ext} + y_{ext} \to C \to G^{\urcorner}$ as depicted below
\[
\scalebox{0.75}{

    \tikzfig{combinatorial_semantics/DPOI_pushout_complement}
}
\]

is a \textit{boundary} complement if $[i_c, i_j]$ is mono 
 and there exist $n_c : n_{ext} \to C$ and $m_c : m_{ext} \to C$ 
 making the above triangle commute and such that
there exists a \textit{not necessarily well-typed} mda-cospan

@@

where $c_i$ and $c_o$ are computed as
 $c_i = g_i \setminus (l_i \setminus x_{ext})$ and
 $c_o = g_o \setminus (l_o \setminus y_{ext})$.

 ALEKSEI: ARE MORPHISMS $c_i + y_{ext} \to C$ AND THE SYMMETRIC ONE OBVIOUS?
\end{definition}
%
%\begin{figure}
%    \centering
%    \[
%    \scalebox{0.5}{\tikzfig{combinatorial_semantics/interfaces_change_example}}\]
%    \caption{Hypothetical DPO square for $\catname{MACsp_{D}(EHyp_{\Sigma})}$}
%    \label{fig:interface_change_example}
%\end{figure}

@@

\begin{remark}
Note that cospan $n_{ext} \xrightarrow{} c_i + y_{ext} \xrightarrow{} C \xleftarrow{} c_o + x_{ext} \xleftarrow{} m_{ext}$ is not necessarily well-typed.
The cospan's input internal interface may contain nodes that previously were a part of the output internal interface and similarly for output internal interface.
\end{remark}

\begin{remark}
    The uniqueness and existence properties of such pushout complement are discussed in the appendix (Proposition~\ref{prop:boundary_complement} and Proposition~\ref{prop:complement_existence}).
\end{remark}

% \begin{proposition}[Proposition~\ref{prop:boundary_complement} restatement]
%     The boundary complement in~\ref{def:boundary_new} when exists is unique.
% \end{proposition}

\begin{definition}[Down-closed subgraph]
    We call a subgraph $\mathcal{H}$ of an e-hypergraph $\mathcal{G}$ \emph{down-closed} if for all $e \in E_{\mathcal{H}}$,   every element of $(e]$ is also in $\mathcal{H}$.
\end{definition}

@@

    \begin{enumerate}
        \item $m$ is monomorphism
        \item $m(L)$ is a convex down-closed subgraph of $G$
    \end{enumerate}
\end{definition}
\

\begin{definition}[Rewriting in $\WellTypedMdaEcospans$]
    Given $G \xleftarrow{} g_i + g_o \xleftarrow{} n_{ext} + m_{ext}$ 
    and $H \xleftarrow{} h_i + h_o \xleftarrow{} n_{ext} + m_{ext}$ in $\Ecospans$ with the outermost interfaces $n_{ext} + m_{ext}$, $G$ rewrites convexly into $H$ with the outermost interface $n_{ext} + m_{ext}$ --- notation $( G \xleftarrow{} n_{ext} + m_{ext} ) \Rrightarrow_{\mathcal{R}}^{+} (E \xleftarrow{} n_{ext} + m_{ext} )$ --- if there exists a rule $L \xleftarrow{} l_i + l_o \xleftarrow{} x_{ext} + y_{ext} \xrightarrow{} r_i + r_o \xrightarrow{} R$ with the outermost interfaces $x_{ext} + y_{ext}$ in the set of rewrite rules $\mathcal{R}$ and object $C$ and cospan arrows $x_{ext} + y_{ext} \xrightarrow{} C \xleftarrow{} n_{ext} + m_{ext}$ such that the diagram below commutes and its marked squares are pushouts

@@

    \begin{enumerate}
        \label{dpoi:assumptions}
        \item $m : L \to G$ is a convex down-closed match;
        \item $x_{ext} + y_{ext} \to C \to G$ is a boundary complement in the leftmost pushout.
    \end{enumerate}
\end{definition}

The internal interfaces of the resulting e-hypergraph $H$ from the cospan
 $n_{ext} \xrightarrow{} h_i \xrightarrow{} H \xleftarrow{} h_o \xleftarrow{} m_{ext}$
 are constructed as $h_i = (r_i \setminus x_{ext}) + c_i$ 
 and
 $h_o = (r_o \setminus y_{ext}) + c_o$.
 $h_i \to H$ is defined as a copairing of 
 $f_1 : r_i \setminus x_{ext} \to R \to H$ 
 and 
 $f_2 : c_i \to C \to H$. 
Since both $f_1$ and $f_2$ are mono, their copairing is also a mono. 
Similarly for $h_o$.

@@

if 
\begin{align*}    
    G \xleftarrow{\langle a_2,b_2 \rangle} g_i + g_{o}& \xleftarrow{\langle a_1, b_1 \rangle} n_{ext} + m_{ext} \\ &\Rrightarrow^{+}\\ H \xleftarrow{\langle a_2,b_2 \rangle} h_i + h_{o}& \xleftarrow{\langle a_1, b_1 \rangle} n_{ext} + m_{ext}
\end{align*}

under a rewrite rule
\[
    \langle l_{i} + l_{o} \xrightarrow{\langle a_2, b_2 \rangle} L \xleftarrow{\langle a_1, b_1 \rangle} x_{ext} + y_{ext},
    r_{i} + r_{o} \xrightarrow{\langle a_2, b_2 \rangle} R \xleftarrow{\langle a_1, b_1 \rangle} x_{ext} + y_{ext} 
    \rangle
\]

@@

\begin{theorem}[Existence of pushouts in $\catname{EHyp_{\Sigma}}$]
\label{th:existence_of_pushouts}
Consider the following span in $\catname{EHyp_{\Sigma}}$
% https://q.uiver.app/#q=WzAsMyxbMCwwLCJaIl0sWzIsMCwiWCJdLFswLDIsIlkiXSxbMCwxLCJmIl0sWzAsMiwiZyIsMl1d
\[\begin{tikzcd}
	Z && X \\
	\\
	Y
	\arrow["f", from=1-1, to=1-3]
	\arrow["g"', from=1-1, to=3-1]
\end{tikzcd}\]
such that
\begin{enumerate}
\label{pushout:assumptions}
    \item $Z$ is a \textit{descrete} e-hypergraph.
    \item $f,g$ are monos.
    \item $[f_{V}(v_i)) = [f_{V}(v_j))$ and $[g_{V}(v_i)) = [g_{V}(v_j))$ for all $v_{i},v_{j}$ in $V_{Z}$.
    \item If $[f_{V}(v)) \not = \varnothing$ then $[g_{V}(v)) = \varnothing$ and if $[g_{V}(v)) \not = \varnothing$ then $[f_{V}(v)) = \varnothing$.
    \item $\consistency^{\mu}(f(v_i)) = \consistency^{\mu}(f(v_j))$ and $\consistency^{\mu}(g(v_i)) = \consistency^{\mu}(g(v_j))$ for all $v_i,v_j$ in $V_{Z}$.
\end{enumerate}    
then the pushout $X +_{f,g} Y$ exists.
\end{theorem}
\begin{proof}
    Recall the diagram below.
    \[
        \begin{tikzcd}
        Z \arrow[r, "f"] \arrow[d, "g"']                                   & X \arrow[d, "\sfrac{i_1}{\sim R}"] \arrow[rdd, "j_1", bend left] &   \\
        Y \arrow[r, "\sfrac{i_2}{\sim R}"] \arrow[rrd, "j_2"', bend right] & \sfrac{X+Y}{\sim R} \arrow[rd, "u"]                              &   \\
                                                                        &                                                                  & Q
        \end{tikzcd}
    \]
    Then, the pushout of e-hypergraphs $X$ and $Y$ is computed in two steps.
    First, a coproduct of $X+Y$ is computed which is 
    \[
        X + Y = \{V_{X} + V_{Y}, E_{X} + E_{Y}, s_{X+Y}, t_{X+Y}, \consistency_{X+Y}, <_{c,X+Y} \}
    \]

@@

\begin{definition}[Definition 3.16~\cite{Frobenius}]
Morphisms $K \xrightarrow{i} L \xrightarrow{m} G$ satisfy the \textit{no-dangling} condition if, for every hyperedge not in the image of $m$, every node of its source and target is either (i) not in the image of $m$ or (ii) in the image of $i ; m$.
They satisfy the no-identification condition if any two nodes merged by $m$ are in the image of $i$.
\end{definition}
